A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-09 tex/codatadp.tex
Reorganize approach for solution soundness proof.  Cleanup.

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}

\def\AIF{\qtab\keyword{if}\ }
\def\ATHEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\AFI{\untab}
\def\ARETURN{\keyword{return}\ }
\def\AENDPROC{\untab}

@@

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In particular, if there is a mapping $\sigma$ over equivalence classes such that all paths $p$ from $t_2$
have a corresponding path $p \sigma$ from $t_1$,
then $t_1$ must be equal to $t_2$.
We demonstrate this rule concretely with the following example.
\rem{ensure that \const{LCons} understood}

@@

\begin{lemma}[Solution Soundness] \label{lem:ss}
If there exists a derivation tree with root node $\Ec$ containing a saturated node $\Fc$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
We assume that:
\begin{enumerate}
\item Each equivalence class of $\Fc$ contains at least one variable,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Nondata$ contains a constant,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Data \cup \Codata$ contains a constructor term.
\end{enumerate}
The first statement comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
The second statement can be assumed for $x : \tau$, since $\tau$ is infinite.
Hence, we can assign constants for each such equivalence class.
Moreover, since $\Fc$ contains $\thD$-constraints only, these constants can be chosen arbitrarily.
The third statement can be assumed by \rem{TODO}.

We construct a model $\M$ of $\thD$ that interprets constructors and correctly-applied selectors in the obvious way,
interprets wrongly-applied selectors based on equality information in $\Fc$,
interprets the variables and $x$ of $\Fc$ based on a valuation $\Val$, which is constructed as follows.
For each equivalence class $e : \tau$, where $\tau \in \Nondata$ containing constant $c$,
$\Val(x) = c$ for each $x \in e$.
For each $e : \tau$, where $\tau \in \Data$ containing the constructor term $C( \vec u )$ where the equivalence classes $\ec{\vec u}$ contains the variables $\vec y$,
$\Val(x) = C( \Val( \vec y ) )$ for each $x \in e$.
Since \rn{Cyclic} does not apply, this definition is well-defined. 
For each $e : \tau$, where $\tau \in \Codata$, 
$\Val( x ) = \interp{ \tpath{\Fc}{e} }$ for each $x \in e$.
Since neither \rn{Cong} or \rn{Bisimilar} applies to $\Fc$, we have that for all variables $x$ and $y$:
\begin{eqnarray} \label{eqn:m-bij-dt}
\Val(x) = \Val(y) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.

If $\tau \in \Nondata$, then $\M( t_1 ) = \M( t_2 )$ by construction.

If $\tau \in \Data$, then assume the equivalence class of $t_1$ and $t_2$ contains variable $x$.
We show by induction on the size of $\Val( x )$ that $\M( t ) = \Val( x )$ for each term $t \in \ec{x}$, which implies $\M \models t_1 \teq t_2$.
In the case where $t$ is a variable, then $\M( t ) = \Val( x )$ since $t$ and $x$ are assigned identical values by construction.
In the case where $t$ is a constructor term $C( \vec u )$, say our construction built a valuation for $x$ based on the constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val( x )$.
In the case where $t$ is a selector term $s^j_k( u )$,
then since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.
Let $x_1, \ldots, x_n$ be the variables in equivalence classes $\ec{s^1_i( u )}, \ldots, \ec{s^{a_i}_i( u )}$.
By our construction, $\M( u ) = C_i( \Val( x_1 ), \ldots, \Val( x_{a_i} ) )$.
If $i = k$, then $\M( t ) = \M( s^j_k( u ) ) = s^j_k( \M( u ) ) = \Val( x_j )$, which is $\Val( x )$ since $x_j \in \ec{s^j_i( u )} = \ec{x}$.
If $i \neq k$, then $\thD$ does not constraint the value of $\M( t )$ and since $s^j_k( \M( u ) )$ is otherwise unconstrained in our model construction since \rn{Cong} does not apply, 
we may interpret $s^j_k( \M( u ) )$ as $\Val( x )$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}

@@

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In this rule, $\sigma$ is a complete mapping over the equivalence classes that occur as sources in the edges of paths in $\tpath{\Ec}{\ec{t_2}}$.
The rule applies if $\tpath{\Ec}{\ec{t_1}}$ is comprised of the paths $p \sigma$ for each (and only) the paths $p \in \tpath{\Ec}{\ec{t_2}}$,
in which case $t_1$ must be equal to $t_2$.
We demonstrate this rule concretely with the following example.
\rem{ensure that \const{LCons} understood}

@@

\begin{lemma}[Solution Soundness] \label{lem:ss}
If there exists a derivation tree with root node $\Ec$ containing a saturated node $\Fc$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
We construct a model $\M$ of $\thD$ that interprets constructors and correctly-applied selectors in the obvious way,
interprets wrongly-applied selectors based on equality information in $\Fc$,
interprets the variables and $x$ of $\Fc$ based on a mapping $\Val$ from the equivalence classes of $\Fc$ to values, 
which is constructed based on the following rules.

\rn{Ord}
\(
\inferrule{
  \ec{x} \not\in \vdom(\Val)
  \quad
  x : \tau
  \quad
  \tau \in \Nondata
  \quad
  c \not\in \vrange(\Val)
}{
  \Val := \Val \cup \ec{x} \mapsto c
}
\)

\rn{D-Ind}
\(
\inferrule{
  \ec{x} \not\in \vdom(\Val)
  \quad
  x : \tau
  \quad
  \tau \in \Data
  \quad
  C( \vec t ) \in \ec{x}
  \quad
  \ec{\vec t} \in \vdom(\Val)
}{
  \Val := \Val \cup \ec{x} \mapsto C( \Val \ec{\vec t}  )
}
\)

\rn{D-Base}
\(
\inferrule{
  \ec{x} \not\in \vdom(\Val)
  \quad
  x : \tau
  \quad
  \tau \in \Data
  \quad
  \ec{x} \text{ ctr-free}
  \quad
  c \not\in \vrange(\Val)
}{
  \Val := \Val \cup \ec{x} \mapsto c
}
\)

We assume these rules are applied in the order in which they are written, for instance, 
the rule \rn{D-Ind} is applied to completion before applying \rn{D-Base}.

\begin{eqnarray} \label{eqn:m-bij-dt}
\Val \ec{x} = \Val \ec{y} \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

\begin{comment}
We assume that:
\begin{enumerate}
\item Each equivalence class of $\Fc$ contains at least one variable,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Nondata$ contains a constant,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Data \cup \Codata$ contains a constructor term.
\end{enumerate}
The first statement comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
The second statement can be assumed for $x : \tau$, since $\tau$ is infinite.
Hence, we can assign constants for each such equivalence class.
Moreover, since $\Fc$ contains $\thD$-constraints only, these constants can be chosen arbitrarily.
The third statement can be assumed by \rem{TODO}.

For each equivalence class $e : \tau$, where $\tau \in \Nondata$ containing constant $c$,
$\Val(x) = c$ for each $x \in e$.
For each $e : \tau$, where $\tau \in \Data$ containing the constructor term $C( \vec u )$ where the equivalence classes $\ec{\vec u}$ contains the variables $\vec y$,
$\Val(x) = C( \Val( \vec y ) )$ for each $x \in e$.
Since \rn{Cyclic} does not apply, this definition is well-defined. 
For each $e : \tau$, where $\tau \in \Codata$, 
$\Val( x ) = \interp{ \tpath{\Fc}{e} }$ for each $x \in e$.
Since neither \rn{Cong} or \rn{Bisimilar} applies to $\Fc$, we have that for all variables $x$ and $y$:
\begin{eqnarray} \label{eqn:m-bij-dt}
\Val(x) = \Val(y) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}
\end{comment}

Without loss of generality, we will assume in the following that all equivalence class of $\Fc$ contain at least one variable,
which comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
We let $\M( x ) = \Val \ec{x}$ for each variable $x$ in $\Fc$.
We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.

If $\tau \in \Nondata$, $t_1$ and $t_2$ must be variables, and thus $\M( t_1 ) = \M( t_2 )$ by construction.

If $\tau \in \Data$, then assume the equivalence class of $t_1$ and $t_2$ contains variable $x$.
We show by induction on the size of $\Val \ec{ x }$ that $\M( t ) = \Val \ec{x}$ for each term $t \in \ec{x}$, which implies $\M \models t_1 \teq t_2$.
In the case where $t$ is a variable, then $\M( t ) = \Val \ec{ t } = \Val \ec{ x }$.
In the case where $t$ is a constructor term $C( \vec u )$, 
then $\Val \ec{x}$ must have been assigned by \rn{D-Ind} for some constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val \ec{ x }$.
In the case where $t$ is a selector term $s^j_k( u )$,
then since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.
Let $x_1, \ldots, x_n$ be the variables in equivalence classes $\ec{s^1_i( u )}, \ldots, \ec{s^{a_i}_i( u )}$.
By our construction, $\M( u ) = C_i( \Val \ec{ x_1 }, \ldots, \Val \ec{ x_{a_i} } )$.
If $i = k$, then $\M( t ) = \M( s^j_k( u ) ) = s^j_k( \M( u ) ) = \Val \ec{x_j}$, which is $\Val( x )$ since $x_j \in \ec{s^j_i( u )} = \ec{x}$.
If $i \neq k$, then $\thD$ does not constraint the value of $\M( t )$ and since $s^j_k( \M( u ) )$ is otherwise unconstrained in our model construction since \rn{Cong} does not apply, 
we may interpret $s^j_k( \M( u ) )$ as $\Val \ec{ x }$.
